### 2016-10-09 thesis_paper/source.tex
automatic triggered changes

@@

\mainmatter
\linespread{1}\selectfont
\chapter{INTRODUCTION} 
\linespread{2}\selectfont\justify
\hspace{3em} In various sciences simulation systems take big place and often times they may be the clue for results. One of such sciences, which is primarily related to this paper, is physics. In this case, from computer engineering point of view, simulations have the following flow:

@@

\noindent\hspace{3em}The purpose of this paper is primarily focused on the third step, which basically involves the entire simulation data to be processed by the analysis software. It is important to underline scale of data being discussed. Molecular simulations represent systems starting from hundreds thousands of atoms and more which are tracked by frames as snapshot of specific time and can last for months. Thus, the size of data starts from couple of gigabytes up to terabytes and petabytes.

\noindent\hspace{3em}Big data processing is becoming one of the key issues with the amount of data being generated by modern systems. Eventually, this data is required to be processed on the fly, thus, analysis software should be able to handle massive data in a very short period of time. When working with huge volume of data, there are non-trivial issues arise. For example, it can take days and weeks to analyze big enough data sets, because the data cannot be simply loaded into memory, since it can get up to terabytes, thus, it has extra overhead because of traditional random access disk I/O framework to read from disk chunks by chunks. Beside it, analysis of the data can get even more complicated, since going through certain parts of the data once would not be enough, which leads to low throughput and efficiency of loaded data. One such example, the data analysis approach has polynomial complexity just for reading the data in order to come up with result, and since the data can't simply be saved in memory, it raises the overhead of disk I/O, too. Pull-based architectures in data processing engines are inefficient, since having a set of specific queries, in order to compute them all, it is needed to fetch data, filter it, and apply needed formulas. It is inefficient, since for every query the same chunk of data needs to be pulled into the memory at least the number of queries times or more in case of more complex queries.

@@

\noindent\hspace{3em}The primary focus and problem in this paper is scientific data analysis. Examples brought up above are real life problems that are very similar to the scientific problem that we are focusing on in terms of data sizes. Particles simulation is one of the most popular methods of analyzing certain chemical reactions, physical processes, or other behavior of different materials.
Molecular simulations (Molecular Dynamics) are applied in different fields and represent a method of analyzing physical movements of particles, atoms, and molecules in a fixed space with a given period of time, apparently with a possibility of giving initial state for each item that is involved in the process and can affect the system. These systems are most of the time N-body simulations. The number of atoms in simulations vary in hundred of thousands, particularly, we may observe two simulation systems of a collagen fiber structure and dipalmitoylphosphatidylcholine (DPPC) bi-layer lipid system consisting of 890,000 and 402,400 atoms as mentioned and illustrated in previous work of the system \cite{mainPaper}. Simulation data represents number of records of physical properties such as mass, charge, velocity, coordinates, and forces for each item aggregated as frames, where each frame represents a snapshot of time, placed with a fixed time interval which may also vary depending on the simulation itself and simulation precision requirement. "Quantities measured during the simulations are analyzed to test the theoretical
model~\cite{Frenkel:api01,Landau:cup05}. In short, the MS is proven
and powerful tool for understanding the inner-workings of a
biological system, by supplying a model description of the
biophysical and biochemical processes that are being unfold at a
nanoscopic scale." \cite{mainPaper}

@@

\noindent\hspace{3em}In this paper, we are incorporating parallelization into the processing part of the proposed design by means of CUDA programming language for GPU. Since storage of the simulation data is very expensive, it might come to the point of analyzing the data on the fly (meaning running the simulation and analyzing it at the same time in a streaming manner), which leads to a problem of optimizing the processing part of the design proposed in the previous paper, because time spent on generating data should be tried to conceal the time spent on the processing part by means of overlapping or simply running it in a quick manner. We will talk about this in details as well in further sections.

\section{Problem Statement}
\hspace{3em} Simulation software systems, in general, follow the same methodology of running and storing simulation data. The simulation software system examples are: Gromacs \cite{GROMACS4}, VMD\cite{VMD}, MDAnalysis~\cite{MDAnalysis}, Wordom~\cite{wordom}, MD-TRACKS~\cite{MDtracks}, SimulaidOne~\cite{Simulaid}, Charmm~\cite{CHARMM}. In the type of simulations brought up as examples above, the flow of the data is the following. Once the simulation is run, the output files are contained as trajectory files with descriptors (they contain information about space dimensions, number of atoms and frames, etc.) that can be easily transposed into simple flat files containing the physical properties atom by atom, frame by frame, which are consequently read and processed by the proposed push-based system. Since we have certain amount of queries needed to be run on given simulation data, generating high I/O traffic followed by design of pull-based system is not considerable. The approach proposed in the previous paper is very good in terms of performance in comparison with the original pull-based system\cite{mainPaper}. As you can see on Figure \ref{fg:sample_estimation_old_paper}, it gets up to over 100 times of speedup which is quite crucial, if we think about days of computations. The problem is still that some of the queries processed by those means are still improvable, especially taking into consideration the fact that in the used previous works for calculation of 2-body functions, which take the biggest time for processing~\cite{ytu:icde09, EDBT12}, we might have some error bounds, which might be unacceptable for certain simulation analysis where sacrifice on loss of data is unbearable. Beside it, although we might have huge performance boost on specific data structure as density map, we still have certain expectation from data nature (such as its uniformity), thus, it makes sense to have desire to simplify the processing and still having pure computation based on any values, considering that the queries should be available to be pre-programmed by user in a separate module of the code. All the benchmarks will be done in comparison with already push-based system versus push-based system with GPU improvements. We think that it is  not fair to compare pull-based against GPU push-based approach, since the goal of the paper is to demonstrate capabilities of GPU approach that can beat simple sequential push-based system.

@@

\noindent\hspace{3em}Taking an advantage of GPU devices nature and incorporating parallelism and streaming for different types of queries, we have come up with a good speed up over sequential processing which gets up to 60 times faster, which is just for sampling with no deep improvements on individual queries (since we believe that it should represent average user usage). Original push-based system already had a good speed up in terms of performance in comparison with original pull-based approach and floats around more than 100 times faster depending on number of atoms, frames, and workload. The module for generation of mock data in Python  was used for benchmarks, and it represents realistic and pessimistic approach of estimation, since there is no data distribution for physical properties values. We believe that it might be quite important, because experiments may vary widely. This mock data has exactly the same information that would come with Gromacs simulation files, and it has exactly the same format as in the previous paper \cite{mainPaper} right before loading it to memory. This improved version has been developed on Amazon Elastic GPU \cite{amazonGPU} nodes that consequently can be replicated and scaled up becoming more of a streaming distributed processing engine (which can be very effective on costs, since it is always easy to scale them up, shut them down and spin them back up), as well as on a simple desktop computer with a GPU device.
In general, major contribution of this paper is the framework of push-based system with an incorporated parallelism based on CUDA programming language. This tool can be integrated into open source project of Gromacs as a custom plugin that can be used for optimized analytic queries to be run over data sets live streaming as well as from files. This will be discussed in more details in further chapters.
The structure of the framework which will be described later is developed in such a way that one can easily add new queries based on their complexity or modify existing ones adding appropriate selections and filters.

@@

\section{Stream Processing}
\hspace{3em} Another major modern issue is live data streaming and live processing, which is also possible with our system. Basically, streaming function is a big challenge and need nowadays, because such networks as Twitter and Facebook are able to keep track of live trends and events, including people's reaction and feedback that have to be reflected real time for tremendous amount of users across the globe. Borealis Stream Processing Software is a tool being developed by MIT, Brown, and Brandeis Universities. It's primary goal is a distributed stream processing and based on Medusa and Aurora. Some of the major functions of the software are highly available distributed large streams processing live, dynamic and highly-scalable with possibility of changing queries live and dynamic vision of query results. \cite{borealis} There is also a similar need in scientific world, since being able to keep track of results in real time might be crucial to be able to make certain explorations.

\noindent\hspace{3em}As a plugin for optimizing computation of scientific queries against large data sets being produced, Push-based System for Molecular Simulation Data Analysis with GPU improvement should be able to handle data streams as well, in order to get real time results and saving up on time, instead of pulling everything from files later on.

@@

\vspace{3em}

@@

\section{One Body Functions}
\hspace{3em} \emph{\textbf{One body functions}} such as sum of masses, center of mass, or simple counting with selection  are of a complexity $O(n)$, thus, in order to get the results for them going through every atom (essentially over the entire data) only once is enough, and taking into consideration the fact that we are going with push-based system, it is done at the same time for all of them.

@@

\section{Multiple Body Functions}
\noindent\hspace{3em}\emph{\textbf{Multiple body functions}} such as SDH and RDF (Spacial Distance Histogram and Radial Distribution function) require computation of distances pairwise across all the particles. This is generally combination of two across N data: 	$C\binom{N}{2}$. These are the most expensive queries, thus, it is important to focus on their implementation taking advantage of GPU nature and its processing power. SDH is a very expensive computation, and we have used some of the existing work of ours\cite{sdhgpu} to incorporate in this design.

@@

\noindent\hspace{3em}There might be some dependencies and preliminary computation possibilities. For example, some queries need total mass, and we could compute it before we push it further, but this can be easily done by user based on the complexity and dependency. It is explained in more details in further sections.

@@

\noindent\hspace{3em}\emph{\textbf{GPU execution}}. From hardware standpoint, it is important to understand that every GPU device consists of SMs (streaming multiprocessors), and each SM has thousands of registers, which can be partitioned among execution threads. Having this in mind, since this is a pure parallel processing power, the goal is to increase hardware throughput and keep highest utilization to exploit every register given. CUDA is an amazing tool that tries to help developers map this hardware capabilities with the actual application layer. In CUDA we have grids, blocks, and threads, where grid is 3 dimensional unit where each cell is a block, and the same relationship applies between blocks and threads\footnote{This explanation doesn't include memory setup in GPU devices, but it is covered in further sections}. Abstract dimensions given by CUDA are not real entities of physical GPU memory and its organization, but it is just a very convenient way of organizing data and tasks to help GPU scheduler pick and execute the instructions. Originally, this mapping was perfect for solving image processing problems or other graphics computation, since dimensions and such division (e.g. by pixels to threads) is very inherent. As a language feature, in CUDA we have to define such a function called 'kernel' function. Essentially, it is the function where we are supposed to map every execution thread with data and perform the actual execution, this approach is widely known as SIMD\footnote{SIMD - Single Instruction Multiple Data}. As a convenience we have separated one-body functions with multiple-body functions into separate kernels.

\noindent\hspace{3em}\emph{\textbf{One-body queries kernel}}.
Commonly used CUDA array reduction\cite{gpureduction} makes the implementation of all the one-body queries obvious. Figure \ref{fg:reduction} represents general idea of the data mapping, where each cell of the vector can be associated with a thread of execution in the kernel. Having an array of atom data structures, we reduce the array accumulating all the needed data from each atom's attributes. We don't expect much performance improvement from coalesced memory access, since the data structure for atom can eliminate any gain of coalesced memory access, though for specific types of queries it could be possible to aggregate certain attributes into merged arrays and take an advantage of it. In this case, the solution should be simple in order to keep it easily extendable for other simple one-body queries for a scientist to program. We fetch all the atoms from GPU global memory, since we use references to atoms once. In this case, allocating shared memory and copying it over would complicate the implementation and possibly create additional overhead for one time reference usage. Finally, memory transportation and data consumption are implemented for user, thus, he only needs to extend the queries module which consequently can be turned into plugins model.

\noindent\hspace{3em}\emph{\textbf{Two-body queries kernel}}.
SDH (or RDF) is the algorithm that requires special treatment in our design. The point, as mentioned earlier, is that this function requires more than one iteration of the given set of atoms, but rather all combinations of pairs in a given frame, thus, it is not linear complexity. An implementation of this type of query might be very non-intuitive because of the nature of GPUs. Another issue is that since to take an advantage on this type of computation, before writing an implementation for it, a user must have a good understanding of GPU architecture which makes the system less user-friendly. At this point, it is more important to achieve good performance improvement results and then take care of ease of system extensions. As we already mentioned, GPU threads execute in parallel and consume certain amount of work followed by best try of all hardware capacities utilization, but, unfortunately, it is also important to take care of equal workload for every thread independently, so, that utilization is kept high. For this reason, we've come up with a solution that should increase performance of the computation which will be described later.

@@

\noindent\hspace{3em}\emph{\textbf{Architecture overview.}}
This section is to emphasize general architecture of the system and summarize the work flow as well as the data flow. In total, we have flat files with ready flat values for each atom aggregated in frames coming into the program as data input using common disk I/O framework of an operating system. Then, once we have loaded a frame (which is considered to be a chunk of the entire data set) onto RAM of our machine, we copy over the entire frame onto the GPU device global memory. It is important that, at this point, we never write any values to simulation files or any other data related to simulation, thus, we don't have to care about any synchronization of incoming data. After this, we have two predefined kernel functions (which were explained before), where first kernel stands for one-body queries, and the second one is for two-body query SDH. Just for architecture clearance and possible future improvements, two kernels are executed in two different streams\footnote{In CUDA stream is another abstract layer of organizing execution threads. They are not guaranteed to be executed at the same time or after each other, GPU scheduling architecture is completely different from CPU having a warp to be single scheduling unit at maximum of 32 threads at a time} (You can see them on Figure\ref{fg:overallarchitecture}). In each kernel we apply certain heuristics to improve memory access using such architecture features of GPU as shared memory and blocking of kernel threads.

\noindent\hspace{3em}\emph{\textbf{GPU grid and block sizes.}} As it has already been mentioned, utilization of hardware on GPU device is important, because it is the key to performance improvement, and the higher utilization is, the bigger improvement is.
The approach we came up with is 1 execution thread per atom. For CUDA version 2.0 maximum number of threads per block is 1024. Blocks are 3 dimensional entities in a 3 dimensional grid. The less abstract unit is, the less overhead we have in communication between them, thus, we try to utilize smallest units more. For example, for 1000,000 atoms, we would allocate 1024 threads per block as a static hard coded value, since we usually have always more than 1024 atoms in simulations. One dimension of grid can take up to 64,000 blocks, thus, if N is bigger than $1024 * 64,000$, than the rest would be allocated to $grid.y$ and $grid.z$ respectively. We have not considered numbers bigger than GPU can allocate, which could be explored in future work. It is important to understand that even though there are abstract entities of grids and blocks, it is not real hardware capacities, but they can affect scheduling and performance consequently. Decision to go with atom to thread execution is because it is pretty convenient to keep high utilization of hardware and still not lose idea if we are about to incorporate more functions, which makes the system more user-friendly, though later on as a plugin we left some room to add data transformer between RAM and GPU global memory, because there are a lot of custom ways to aggregate data and take an advantage of coalesced memory access.

@@

\section{Computation Work Flow}

@@

\subsection{One-body Functions Implementation}

@@

\subsubsection{One-body Functions Implementation In CPU Version}

\hspace{3em} Basically being given a vector of atoms in CPU version implementation (the one that we use for speedup benchmark) we loop through every atom and accumulate all the needed information to RAM. Once we have the result for one-body functions, we take it out for further usage, it can as well be used for further calculations where there are dependencies on such things as total mass, total charge, etc.

@@

\subsubsection{One-body Functions Implementation In GPU Version}

\hspace{3em} The approach in GPU version of one-body computation code is a lot different, because there are two major changes: the way we deal with the data in memory and mapping of GPU thread kernels to atoms. Both of these approaches drastically change the way we perceive the solution implementation. Basically in the kernel function we assume we've pre-loaded all the needed atoms for current calculations (loading into memory and higher level parts will be described in later sections).

@@

\State $\textit{sharedResult} \gets \textit{pointer to shared memory head (in this case declared element size as float)}$

@@

\State
\State $\textit{// Here we will starting collecting information from each block which will be done by first thread of each block}$
\If {$\textit{threadIdx.x} = 0$}
\State  $\textit{atomicAdd(\&results->mass, sharedResult[0]);}$ \Comment{All atomicAdd functions here are GPU atomic}
\State  $\textit{atomicAdd(\&results->charge, sharedResult[1]);}$
\State  $\textit{atomicAdd(\&results->inertiaX, sharedResult[2]);}$
\State  $\textit{atomicAdd(\&results->inertiaY, sharedResult[3]);}$
\State  $\textit{atomicAdd(\&results->inertiaZ, sharedResult[4]);}$
\State  $\textit{// other functions to accumulate from block shared memory...}$
\EndIf

@@

\subsection{Two-body Functions Implementation}

\hspace{3em} Since one of the physical queries is RDF which has underlying SDH calculation, we will describe our approach with SDH for CPU and GPU versions.

\subsubsection{Two-body Functions Implementation In CPU Version}

\hspace{3em} Implementation for CPU version of SDH is very simple, we do 2 nested loops that calculate histogram. Since SDH criteria and parameters might vary by bucket width and box size for the simulation space, we've used one sample size for most of them, though results don't change a lot for sequential computation.

@@

\subsubsection{Two-body Functions Implementation In GPU Version}

\hspace{3em} The approach of solving SDH with GPU is absolutely different just like with one-body queries, though number of computations are the same. We do brute force  $O(\frac{N^2}{2})$ complex solution which is pretty much looking at distance between every pair of atoms. In Pseudocode \ref{alg:sdhgpu} we have tried to provide main logic flow that explains how exactly we implemented kernel for SDH computation. 
Once the first kernel is finished we run the two-body kernel with its own stream. As it has already been mentioned, the bigger workload is in this kernel, because it has higher complexity. GPU streaming multiprocessors are very powerful for straightforward primitive independent calculations, and each core is much weaker than a moden CPU which has a whole bunch of L caches with some prediction behavior systems for code divergence. As you can see on Figure \ref{fg:threadload}, unbalanced thread load assumes for GPU kernel to check every time if the target atom can interact with other atom and if it has to wait till other threads finish their jobs. Thus, we had to come up with a strategy of making equally balanced thread jobs and distribute them among the threads leaving atom to thread mapping as is.

@@

\subsection{Data Flow Implementation In CPU}

\hspace{3em} As we have described in earlier sections, our CPU is primarily used for data flow management, triggering of GPU functions, data structure changes, and GPU computation stream manipulations.
%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%      CPU FLOW %%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{algorithm}
\caption{Main algorithm in CPU for managing data and pushing to GPU}
\label{alg:cpuflow}
\begin{algorithmic}[1]
\Procedure{runsinglekernel}{atomsCount, atoms, workload}

@@

\subsection{GPU}

>>>>

\usepackage{float}
\floatstyle{plaintop}
\restylefloat{table}

@@

\mainmatter
\linespread{1}\selectfont
\chapter{INTRODUCTION} 
\linespread{2}\selectfont\justify
\hspace{3em} In various sciences simulation systems are very crucial and might be the key factors for results. One of such sciences, which is primarily related to this paper, is physics. In this case, from computer engineering point of view, simulations have the following flow:

@@

\noindent\hspace{3em}The purpose of this paper is primarily focused on the third step, which basically involves the entire simulation data to be processed by the analysis software. It is important to underline scale of data being discussed. Molecular simulations represent systems starting from thousands of atoms and more which are tracked by frames as snapshot of specific time and can last for months. Thus, the size of data starts from couple of gigabytes up to terabytes and petabytes.

\noindent\hspace{3em}Big data processing is becoming one of the key issues with the amount of data being generated by modern systems. Eventually, this data is required to be processed on the fly, thus, analysis software should be able to handle massive data in a very short period of time. When working with huge volume of data, there can be additional issues while using regular tools. For example, it can take days and weeks to analyze big enough data sets, because the data cannot be simply loaded into memory, since it can get up to terabytes, thus, it has extra overhead because of traditional random access disk I/O framework to read from disk chunks by chunks. Beside it, analysis of the data can get even more complicated, since going through certain parts of the data once would not be enough, which leads to low throughput and efficiency of loaded data. One such example, the data analysis approach has polynomial complexity just for reading the data in order to come up with result, and since the data can't simply be saved in memory, it raises the overhead of disk I/O, too. Pull-based architectures in data processing engines are inefficient, since having a set of specific queries, in order to compute them all, it is needed to fetch data, filter it, and apply needed formulas. It is inefficient, since for every query the same chunk of data needs to be pulled into the memory at least the number of queries times or more in case of more complex queries.

@@

\noindent\hspace{3em}The primary focus and problem in this paper is scientific data analysis. Examples brought up above are real life problems that are very similar to the scientific problem that we are focusing on in terms of data sizes. Particles simulation is one of the most popular methods of analyzing certain chemical reactions, physical processes, or other behavior of different materials.
Molecular simulations (Molecular Dynamics) are applied in different fields and represent a method of analyzing physical movements of particles, atoms, and molecules in a fixed space with a given period of time, apparently with a possibility of giving initial state for each item that is involved in the process and can affect the system. These systems are most of the time N-body simulations. The number of atoms in simulations vary in hundred of thousands, particularly, we may observe two simulation systems of a collagen fiber structure and dipalmitoylphosphatidylcholine (DPPC) bi-layer lipid system consisting of 890,000 and 402,400 atoms as mentioned and illustrated in previous work of the system \cite{mainPaper}. Simulation data represents number of records of physical properties such as mass, charge, velocity, coordinates, and forces for each item aggregated as frames, where each frame represents a snapshot of time, placed with a fixed time interval which may also vary depending on the simulation itself and simulation precision requirement. "Quantities measured during the simulations are analyzed to test the theoretical
model~\cite{Frenkel:api01,Landau:cup05}. In short, the MS is proven
and powerful tool for understanding the inner-workings of a
biological system, by supplying a model description of the
biophysical and biochemical processes that are being unfold at a
nanoscopic scale." \cite{mainPaper}
\clearpage

@@

\noindent\hspace{3em}In this paper, we are incorporating parallelization into the processing part of the proposed design by means of CUDA programming language for GPU. Since storage of the simulation data is very expensive, it might come to the point of analyzing the data on the fly (meaning running the simulation and analyzing it at the same time in a streaming manner), which leads to a problem of optimizing the processing part of the design proposed in the previous paper, because time spent on generating data should be tried to conceal the time spent on the processing part by means of overlapping or simply running it in a quick manner. We will talk about this in detail as well in further sections.

\section{Problem Statement}
\hspace{3em} Simulation software systems, in general, follow the same methodology of running and storing simulation data. The simulation software system examples are: Gromacs \cite{GROMACS4}, VMD\cite{VMD}, MDAnalysis~\cite{MDAnalysis}, Wordom~\cite{wordom}, MD-TRACKS~\cite{MDtracks}, SimulaidOne~\cite{Simulaid}, Charmm~\cite{CHARMM}. In the type of simulations brought up as examples above, the flow of the data is the following. Once the simulation is run, the output files are contained as trajectory files with descriptors (they contain information about space dimensions, number of atoms and frames, etc.) that can be easily transposed into simple flat files containing the physical properties atom by atom, frame by frame, which are consequently read and processed by the proposed push-based system. Since we have certain amount of queries needed to be run on given simulation data, generating high I/O traffic followed by design of pull-based system is not considerable. The approach proposed in the previous paper is very good in terms of performance in comparison with the original pull-based system\cite{mainPaper}. It gets up to over 100 times of speedup which is quite crucial, if we think about days of computations. The problem is still that some of the queries processed by those means are still improvable, especially taking into consideration the fact that in the used previous works for calculation of 2-body functions, which take the biggest time for processing~\cite{ytu:icde09, EDBT12}, we might have some error bounds, which might be unacceptable for certain simulation analysis where sacrifice on loss of data is unbearable. Beside it, although we might have huge performance boost on specific data structure as density map, we still have certain expectation from data nature (such as its uniformity), thus, it makes sense to have desire to simplify the processing and still having pure computation based on any values, considering that the queries should be available to be pre-programmed by user in a separate module of the code. All the benchmarks will be done in comparison with already push-based system versus push-based system with GPU improvements. We think that it is  not fair to compare pull-based against GPU push-based approach, since the goal of the paper is to demonstrate capabilities of GPU approach that can beat simple sequential push-based system.

@@

\noindent\hspace{3em}Taking an advantage of GPU devices nature and incorporating parallelism and streaming for different types of queries, we have come up with a good speed up over sequential processing which gets up to 60 times faster, which is just for sampling with no deep improvements on individual queries (since we believe that it should represent average user usage). Original push-based system already had a good speed up in terms of performance in comparison with original pull-based approach and floats around more than 100 times faster depending on number of atoms, frames, and workload. The module for generation of mock data in Python  was used for benchmarks, and it represents realistic and pessimistic approach of estimation, since there is no data distribution for physical properties values. We believe that it might be quite important, because experiments may vary widely. This mock data has exactly the same information that would come with Gromacs simulation files, and it has exactly the same format as in the previous paper \cite{mainPaper} right before loading it to memory. This improved version has been developed on Amazon Elastic GPU \cite{amazonGPU} nodes that consequently can be replicated and scaled up becoming more of a streaming distributed processing engine (which can be very effective on costs, since it is always easy to scale them up, shut them down and spin them back up), as well as on a simple desktop computer with a GPU device.
In general, major contribution of this paper is the framework of push-based system with an incorporated parallelism based on CUDA programming language. This tool can be integrated into open source project of Gromacs as a custom plugin that can be used for optimized analytic queries to be run over data sets live streaming as well as from files. This will be discussed in more detail in further chapters.
The structure of the framework which will be described later is developed in such a way that one can easily add new queries based on their complexity or modify existing ones adding appropriate selections and filters.

@@

\section{Stream Processing}
\hspace{3em} Another major modern issue is live data streaming and live processing, which is also possible with our system. Basically, streaming function is a big challenge and need nowadays, because such networks as Twitter and Facebook are able to keep track of live trends and events, including people's reaction and feedback that have to be reflected real time for tremendous amount of users across the globe. Borealis Stream Processing Software is a tool being developed by MIT, Brown, and Brandeis Universities. Its primary goal is a distributed stream processing and based on Medusa and Aurora. Some of the major functions of the software are highly available distributed large streams processing live, dynamic and highly-scalable with possibility of changing queries live and dynamic vision of query results. \cite{borealis} There is also a similar need in scientific world, since being able to keep track of results in real time might be crucial to be able to make certain explorations.

\noindent\hspace{3em}As a plugin for optimizing computation of scientific queries against large data sets, an improved Push-based System for MS improvement should be able to handle data streams as well, in order to get real time results and saving up on time, instead of pulling everything from files later on.

@@

@@

\section{One Body Functions}
\hspace{3em} One body functions such as sum of masses, center of mass, or simple counting with selection  are of a complexity $O(n)$, thus, in order to get the results for them going through every atom (essentially over the entire data) only once is enough, and taking into consideration the fact that we are going with push-based system, it is done at the same time for all of them.

@@

\section{Multiple Body Functions}
\noindent\hspace{3em}Multiple body functions such as SDH and RDF (Spacial Distance Histogram and Radial Distribution function) require computation of distances pairwise across all the particles. This is generally combination of two across N data: 	$C\binom{N}{2}$. These are the most expensive queries, thus, it is important to focus on their implementation taking advantage of GPU nature and its processing power. SDH is a very expensive computation, and we have used some of the existing work of ours\cite{sdhgpu} to incorporate in this design.

@@

\noindent\hspace{3em}There might be some dependencies and preliminary computation possibilities. For example, some queries need total mass, and we could compute it before we push it further, but this can be easily done by user based on the complexity and dependency. It is explained in more detail in further sections.

@@

\noindent\hspace{3em} From hardware standpoint, it is important to understand that every GPU device consists of SMs (streaming multiprocessors), and each SM has thousands of registers, which can be partitioned among execution threads. Having this in mind, since this is a pure parallel processing power, the goal is to increase hardware throughput and keep highest utilization to exploit every register given. CUDA is an amazing tool that tries to help developers map this hardware capabilities with the actual application layer. In CUDA we have grids, blocks, and threads, where grid is 3 dimensional unit where each cell is a block, and the same relationship applies between blocks and threads\footnote{This explanation doesn't include memory setup in GPU devices, but it is covered in further sections}. Abstract dimensions given by CUDA are not real entities of physical GPU memory and its organization, but it is just a very convenient way of organizing data and tasks to help GPU scheduler pick and execute the instructions. Originally, this mapping was perfect for solving image processing problems or other graphics computation, since dimensions and such division (e.g. by pixels to threads) is very inherent. As a language feature, in CUDA we have to define such a function called 'kernel' function. Essentially, it is the function where we are supposed to map every execution thread with data and perform the actual execution, this approach is widely known as SIMD\footnote{SIMD - Single Instruction Multiple Data}. As a convenience we have separated one-body functions with multiple-body functions into separate kernels.

\noindent\hspace{3em}Commonly used CUDA array reduction\cite{gpureduction} makes the implementation of all the one-body queries obvious. Figure \ref{fg:reduction} represents general idea of the data mapping, where each cell of the vector can be associated with a thread of execution in the kernel. Having an array of atom data structures, we reduce the array accumulating all the needed data from each atom's attributes. We don't expect much performance improvement from coalesced memory access, since the data structure for atom can eliminate any gain of coalesced memory access, though for specific types of queries it could be possible to aggregate certain attributes into merged arrays and take an advantage of it. In this case, the solution should be simple in order to keep it easily extendable for other simple one-body queries for a scientist to program. We fetch all the atoms from GPU global memory, since we use references to atoms once. In this case, allocating shared memory and copying it over would complicate the implementation and possibly create additional overhead for one time reference usage. Finally, memory transportation and data consumption are implemented for user, thus, he only needs to extend the queries module which consequently can be turned into plugins model.

\noindent\hspace{3em}SDH (or RDF) is the algorithm that requires special treatment in our design. The point, as mentioned earlier, is that this function requires more than one iteration of the given set of atoms, but rather all combinations of pairs in a given frame, thus, it is not linear complexity. An implementation of this type of query might be very non-intuitive because of the nature of GPUs. Another issue is that since to take an advantage on this type of computation, before writing an implementation for it, a user must have a good understanding of GPU architecture which makes the system less user-friendly. At this point, it is more important to achieve good performance improvement results and then take care of ease of system extensions. As we already mentioned, GPU threads execute in parallel and consume certain amount of work followed by best try of all hardware capacities utilization, but, unfortunately, it is also important to take care of equal workload for every thread independently, so, that utilization is kept high. For this reason, we've come up with a solution that should increase performance of the computation which will be described later.

@@

\noindent\hspace{3em}This section is to emphasize general architecture of the system and summarize the work flow as well as the data flow. In total, we have flat files with ready flat values for each atom aggregated in frames coming into the program as data input using common disk I/O framework of an operating system. Then, once we have loaded a frame (which is considered to be a chunk of the entire data set) onto RAM of our machine, we copy over the entire frame onto the GPU device global memory. It is important that, at this point, we never write any values to simulation files or any other data related to simulation, thus, we don't have to care about any synchronization of incoming data. After this, we have two predefined kernel functions (which were explained before), where first kernel stands for one-body queries, and the second one is for two-body query SDH. Just for architecture clearance and possible future improvements, two kernels are executed in two different streams\footnote{In CUDA stream is another abstract layer of organizing execution threads. They are not guaranteed to be executed at the same time or after each other, GPU scheduling architecture is completely different from CPU having a warp to be single scheduling unit at maximum of 32 threads at a time} (You can see them on Figure \ref{fg:overallarchitecture}). In each kernel we apply certain heuristics to improve memory access using such architecture features of GPU as shared memory and blocking of kernel threads.

\noindent\hspace{3em}GPU grid and block sizes. As it has already been mentioned, utilization of hardware on GPU device is important, because it is the key to performance improvement, and the higher utilization is, the bigger improvement is.
The approach we came up with is 1 execution thread per atom. For CUDA version 2.0 maximum number of threads per block is 1024. Blocks are 3 dimensional entities in a 3 dimensional grid. The less abstract unit is, the less overhead we have in communication between them, thus, we try to utilize smallest units more. For example, for 1000,000 atoms, we would allocate 1024 threads per block as a static hard coded value, since we usually have always more than 1024 atoms in simulations. One dimension of grid can take up to 64,000 blocks, thus, if N is bigger than $1024 * 64,000$, than the rest would be allocated to $grid.y$ and $grid.z$ respectively. We have not considered numbers bigger than GPU can allocate, which could be explored in future work. It is important to understand that even though there are abstract entities of grids and blocks, it is not real hardware capacities, but they can affect scheduling and performance consequently. Decision to go with atom to thread execution is because it is pretty convenient to keep high utilization of hardware and still not lose idea if we are about to incorporate more functions, which makes the system more user-friendly, though later on as a plugin we left some room to add data transformer between RAM and GPU global memory, because there are a lot of custom ways to aggregate data and take an advantage of coalesced memory access.

@@

@@

@@

\noindent\hspace{3em} Basically being given a vector of atoms in CPU version implementation (the one that we use for speedup benchmark) we loop through every atom and accumulate all the needed information to RAM. Once we have the result for one-body functions, we take it out for further usage, it can as well be used for further calculations where there are dependencies on such things as total mass, total charge, etc.

@@

\noindent\hspace{3em} The approach in GPU version of one-body computation code is a lot different, because there are two major changes: the way we deal with the data in memory and mapping of GPU thread kernels to atoms. Both of these approaches drastically change the way we perceive the solution implementation. Basically in the kernel function we assume we've pre-loaded all the needed atoms for current calculations (loading into memory and higher level parts will be described in later sections).

@@

\State $\textit{sharedResult} \gets \textit{pointer to shared memory head}$

@@

\State
\State $\textit{// Here we will starting collecting information from each block}$
\If {$\textit{threadIdx.x} = 0$}
\State  $\textit{atomicAdd(\&results->mass, sharedResult[0]);}$ \Comment{All atomic functions here}
\State  $\textit{atomicAdd(\&results->charge, sharedResult[1]);}$
\State  $\textit{atomicAdd(\&results->inertiaX, sharedResult[2]);}$
\State  $\textit{atomicAdd(\&results->inertiaY, sharedResult[3]);}$
\State  $\textit{atomicAdd(\&results->inertiaZ, sharedResult[4]);}$
\State  $\textit{// other functions to accumulate from block shared memory...}$
\EndIf

@@

\noindent\hspace{3em} Since one of the physical queries is RDF which has underlying SDH calculation, we will describe our approach with SDH for CPU and GPU versions.

\noindent\hspace{3em} Implementation for CPU version of SDH is very simple, we do 2 nested loops that calculate histogram. Since SDH criteria and parameters might vary by bucket width and box size for the simulation space, we've used one sample size for most of them, though results don't change a lot for sequential computation.

@@

\noindent\hspace{3em} The approach of solving SDH with GPU is absolutely different just like with one-body queries, though number of computations are the same. We do brute force  $O(\frac{N^2}{2})$ complex solution which is pretty much looking at distance between every pair of atoms. In Pseudocode \ref{alg:sdhgpu} we have tried to provide main logic flow that explains how exactly we implemented kernel for SDH computation. 
Once the first kernel is finished we run the two-body kernel with its own stream. As it has already been mentioned, the bigger workload is in this kernel, because it has higher complexity. GPU streaming multiprocessors are very powerful for straightforward primitive independent calculations, and each core is much weaker than a moden CPU which has a whole bunch of L caches with some prediction behavior systems for code divergence. As you can see on Figure \ref{fg:threadload}, unbalanced thread load assumes for GPU kernel to check every time if the target atom can interact with other atom and if it has to wait till other threads finish their jobs. Thus, we had to come up with a strategy of making equally balanced thread jobs and distribute them among the threads leaving atom to thread mapping as is.

@@

\noindent\hspace{3em} As we have described in earlier sections, our CPU is primarily used for data flow management, triggering of GPU functions, data structure changes, and GPU computation stream manipulations.
%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%      CPU FLOW %%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{algorithm}
\caption{Main algorithm in CPU for managing data and pushing to GPU}
\label{alg:cpuflow}
\begin{algorithmic}[1]
\Procedure{runsinglekernel}{atomsCount, atoms, workload}

@@

\noindent\hspace{3em} The same CPU has been used for sequential as well as for improved GPU version during experiments with benchmarks. So, for CPU in GPU improved version refer to the same Table \ref{tb:cpuhardware}. For recall, CPU in improved version is used in-between memory loading from file and uploading it to GPU, printing the results to the user, which is not big load. Consequently, we also have GPU device. Hardware summary for GPU device is described in Table \ref{tb:gpuhardware}
